---
title: "Template RMarkdown File"
author: "Robert J. Dellinger"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 12pt  # Set font size to 12pt for readability
output:
  pdf_document: # PDF output
    toc: true  # Table of contents
    toc_depth: 3  # Set TOC depth
    number_sections: false  # section numbering
    fig_caption: true  # Enable figure captions
    fig_width: 6  # Set figure width
    fig_height: 4  # Set figure height
    keep_tex: true
    latex_engine: xelatex #better font and unicode support
bibliography: "Docs/References/references.bib"  
csl: "Docs/References/Styles/apa.csl"  
header-includes:
  # Set page margins and use double-spacing 
  - \usepackage{geometry}
  - \geometry{left=1in, right=1in, top=1in, bottom=1in}
  - \setlength{\textwidth}{6in}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{titlesec}
  - \titleformat{\section}[block]{\normalfont\Large\bfseries\centering}{\thesection}{1em}{}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \setlength{\headheight}{15pt}
  - \fancyhead[L]{}  # Left header (empty)
  - \fancyhead[C]{Template RMarkdown File} 
  - \fancyhead[R]{}  # Right header (empty)
  - \fancyfoot[L]{}  # Left footer (empty)
  - \fancyfoot[C]{\thepage}  # Centered page number
  - \fancyfoot[R]{}  
  - \usepackage{hyperref}  # For clickable links in the PDF
  - \usepackage{amsmath}  # For better math support
  - \usepackage{amsfonts}  # For fonts related to mathematical symbols
  - \usepackage{amssymb}  # For additional math symbols
  - \usepackage{enumitem}  # For better control over list formatting
  - \usepackage{microtype}  # For better typography 
  - \usepackage{fvextra}  # For including graphics
  - \usepackage{fontspec}
  - \setmonofont{Source Code Pro}
editor_options: 
  chunk_output_type: console
---
\newpage


# Introduction

This document outlines the methodology for data cleaning, exploration, and visualization. It is structured to ensure transparency and reproducibility of all analyses.

## Methodology

Briefly describe the methods used in the project, including data sources, cleaning steps, and techniques applied to handle missing or inconsistent data.

### Math Equations with LaTeX

Here is an inline equation: \( E = mc^2 \).

And here is a displayed equation:

\[
f(x) = \int_{-\infty}^{\infty} e^{-x^2} \, dx
\]

Both inline and block math can be rendered seamlessly with LaTeX.

You can also create multiline equations with alignment:

\[
\begin{aligned}
a &= b + c \\
d &= e + f
\end{aligned}
\]

## Setup Packages

The first step in any data analysis is to load the data, options, and packages. This section installs and loads the necessary packages, sets up the output paths, and defines chunk options for the RMarkdown document. 

```{r setup-packages, include=FALSE}

# Set up the environment and clear the workspace
rm(list = ls())

# Install pacman as package manager to automatically install/update packages
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
library(pacman)

# Load/install essential packages
p_load(# packages for data manipulation and wrangling
    tidyverse, stringr, readr, readxl, dplyr, tidyr, janitor, lubridate, glue, purrr, here,
    # packages for data visualization and reporting
    ggplot2, gganimate, ggrepel, ggtext, ggsci, gridExtra, scales, 
    # packages for spatial data analysis and visualization
    sf, tigris, raster, ggmap,
    # packages for statistical analysis
    knitr, rmarkdown, sessioninfo, 
    # packages for formatting, reproducibility and version control
    devtools, remotes, knitcitations, tinytex, formatR
)

# Define output + cache paths
output_path_images <- here("Output", "Figures")
output_path_tables <- here("Output", "Tables")

# Chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  results = 'markup',  # Display text output as code block
  fig.path = paste0(output_path_images, "/Fig-"),
  fig.align = "center",
  out.width = "80%",
  dpi = 800,
  list(width.cutoff = 30), tidy = TRUE)

```

```{r wrap-hook}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```

## Loading Data

The following code loads the data from various sources, including CSV files, Excel files, and shapefiles. The `here` package is used to create file paths that are relative to the project directory, ensuring that the code is portable and reproducible.


```{r load-data}

# Example of loading data from a CSV file
# raw_data <- read_csv(here("Data", "Raw", "data_file.csv"))
# raw_data <- read_excel(here("Data", "Raw", "data_file.xlsx"))
# raw_data <- read_sf(here("Data", "Raw", "data_file.shp"))

```

## Cleaning Data 

The data cleaning process involves several steps to ensure the data is in a suitable format for analysis. This includes handling missing values, correcting data types, and removing duplicates.

```{r wrangle-data}

# cleaning the data using dplyr and tidyr
# cleaned_data <- raw_data %>%
#   clean_names() %>%
#   mutate(column_name = as_factor(column_name)) %>% 
#   mutate(date_column = as.Date(date_column, format = "%Y-%m-%d")) %>% 
#   mutate(numeric_column = as.numeric(numeric_column)) %>% 
#   mutate(accross(everything(), ~str_squish(.))) %>% =
#   drop_na()

```

## Data Exploration

Data exploration is a crucial step in understanding the dataset and identifying patterns or anomalies. This section includes summary statistics, visualizations, and any other relevant analyses to gain insights into the data.


```{r data-exploration}

# Explore the cleaned data using basic summaries:
# glimpse(cleaned_data)
# summary(cleaned_data)
# str(cleaned_data)

```

## Data Visualization

Data visualization is an essential part of data analysis, allowing for the communication of findings in a clear and effective manner. This section includes various plots and charts to illustrate key insights from the data.


```{r data-visualization-table, fig.cap="Example of a scatter plot"}

# Example of creating a summary table 
# summary_table <- cleaned_data %>%
#   group_by(group_var) %>%
#   summarise(mean_value = mean(value_var, na.rm = TRUE)) %>%
#   ungroup() %>%
#   kable() %>%
#   kable_styling(full_width = F, position = "left") 

# save_kable(summary_table, file = here(output_path_tables, "summary_table.html"),
#   bootstrap_options = c("striped", "hover", "condensed"),
#   full_width = F, position = "left")


```


```{r data-visualization-plot, fig.cap="Example of a scatter plot"}

# Example of visualization plot
data("diamonds")

# First Plot (p1) - Scatter plot with smoothing line
p1 <- ggplot(
  subset(diamonds, carat >= 2.2),
  aes(x = table, y = price, colour = cut)
) +
  geom_point(alpha = 0.7) +  
  geom_smooth(method = "loess", alpha = 0.05,
  linewidth = 1, span = 1, formula = y ~ x) +  # Smoothing line
  theme_bw() +  # White background theme
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
    legend.position = "top",  # Position the legend at the top
    text = element_text(size = 10)  # Set text size
  )

# Second Plot (p2) - Histogram of depth for specific carat ranges
p2 <- ggplot(
  subset(diamonds, carat > 2.2 & depth > 55 & depth < 70),
  aes(x = depth, fill = cut)
) +
  geom_histogram(colour = "black",
  binwidth = 1, position = "dodge") +  # Dodged histogram
  theme_bw() +  # White background theme
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
    legend.position = "top",  # Position the legend at the top
    text = element_text(size = 10)  # Set text size
  )

p1_npg <- p1 + scale_color_npg()
p2_npg <- p2 + scale_fill_npg()
grid.arrange(p1_npg, p2_npg, ncol = 2)

# Example of saving a plot
# ggsave(filename = here(output_path_images, "plot_name.png"), 
# plot = last_plot(), width = 6, height = 4)

```

# Literature Cited

Citing the packages and data used in the analysis is important for reproducibility and transparency. The following code generates a bibliography of all loaded packages. Items can be cited directly within the documentation using the syntax `@key` where key is the citation key in the first line of the entry, e.g., @R-base, @R-ggplot2, @R-tidyverse, @R-here. To put citations in parentheses, use `[@key]` instead. 


```{r literature-cited, echo=FALSE, include=FALSE}

references_path <- here("Scripts", "Docs", "References")

# Site packages and data used here 
# Generate citations for loaded packages
knitr::write_bib(.packages(), here(references_path, "references.bib"))
bib_file <- here(references_path, "references.bib")
# Read the bib file
bib <- readLines(bib_file)

# Generate a session info report
sessioninfo <- sessioninfo::session_info()
#print(sessioninfo[1]$platform %>% as_tibble())

# Save session info to a CSV file
write.csv(sessioninfo[1]$platform %>% as_tibble(), here(references_path, "session_info.csv"))
write.csv(sessioninfo[2]$package %>% as_tibble(), here(references_path, "session_info_packages.csv"))

```


