---
title: "**USEPA CDC Social Vulnerability Index (SVI) Analysis**"
author: "Robert J. Dellinger"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    fig_caption: true
    fig_width: 9
    fig_height: 6
    theme: flatly
    highlight: tango
    df_print: paged
csl: "Docs/References/Styles/apa.csl"
bibliography: "Docs/References/references.bib"
---

```{r setup-packages, include=FALSE}

# Clear the workspace
rm(list = ls())

# Install pacman if needed
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
library(pacman)

# Load / install all required packages
p_load(
  # Data manipulation
  tidyverse, dplyr, tidyr, stringr, readr, janitor, glue, purrr, here,
  # Spatial data
  sf, tigris,
  # Visualization
  ggplot2, ggrepel, scales, viridis, gridExtra, RColorBrewer,
  # Tables
  knitr, kableExtra,
  # Reproducibility
  sessioninfo
)

# Suppress tigris tile-cache messages
options(tigris_use_cache = TRUE)

# Output paths
output_path_images <- here("Output", "Figures")
output_path_tables <- here("Output", "Tables")

# Global chunk defaults
knitr::opts_chunk$set(
  echo    = TRUE,
  cache   = FALSE,
  message = FALSE,
  warning = FALSE,
  out.width = "90%",
  fig.path  = paste0(output_path_images, "/SVI-"),
  fig.align = "center",
  dpi = 300
)

```

---

# Introduction

The **CDC/ATSDR Social Vulnerability Index (SVI)** is a publicly available dataset
developed by the Centers for Disease Control and Prevention (CDC) and the Agency for
Toxic Substances and Disease Registry (ATSDR). It uses 16 U.S. Census variables to
rank counties and census tracts by their social vulnerability — the potential negative
effects on communities caused by external stresses on human health.

The SVI is widely used by the **U.S. Environmental Protection Agency (USEPA)** in
environmental justice (EJ) analyses to identify communities that may be
disproportionately burdened by environmental hazards and that may have limited
capacity to prepare, respond to, and recover from those hazards.

## SVI Themes

The 16 variables are organized into **four themes**, each producing a percentile
ranking (RPL). The four theme rankings are combined into a single
**overall SVI score** (`RPL_THEMES`), ranging from 0 (least vulnerable) to
1 (most vulnerable).

| Theme | Key Variables |
|----------|--------------|
| **Theme 1 — Socioeconomic Status** | Poverty, unemployment, housing cost burden, no HS diploma, no health insurance |
| **Theme 2 — Household Characteristics** | Age ≥65, Age ≤17, disability, single-parent households, limited English |
| **Theme 3 — Racial & Ethnic Minority Status** | Racial & ethnic minority population |
| **Theme 4 — Housing Type & Transportation** | Multi-unit structures, mobile homes, crowding, no vehicle, group quarters |

---

# Data Loading

## Download CDC SVI County-Level Data (2022)

The 2022 SVI data at the county level is downloaded directly from the
CDC's public download portal. County-level data covers all 50 states,
DC, and US territories (~3,200 records).

```{r load-svi-data}

# Direct CDC SVI 2022 county-level CSV URL
svi_url <- paste0(
  "https://svi.cdc.gov/Documents/Data/2022/csv/",
  "SVI_2022_US_county.csv"
)

# Cache locally so we do not re-download on every knit
svi_cache <- here("Data", "SVI_2022_US_county.csv")

if (!file.exists(svi_cache)) {
  download.file(svi_url, destfile = svi_cache, mode = "wb", quiet = TRUE)
}

svi_raw <- read_csv(svi_cache, show_col_types = FALSE)

glimpse(svi_raw)

```

---

# Data Cleaning

## Standardise Column Names and Recode Missing Values

The CDC SVI dataset uses `-999` as a sentinel for missing data. We replace
these with `NA` across all numeric columns, clean column names with `janitor`,
and select the key variables needed for analysis.

```{r clean-svi-data}

svi_clean <- svi_raw %>%
  clean_names() %>%
  # Replace CDC sentinel -999 with NA across all numeric columns
  mutate(across(where(is.numeric), ~ na_if(., -999))) %>%
  # Keep identifiers, population, overall SVI, and four theme RPLs
  select(
    fips       = fips,
    state      = state,
    county     = county,
    location   = location,
    area_sqmi  = area_sqmi,
    e_totpop   = e_totpop,
    # Overall SVI percentile ranking (0 = least, 1 = most vulnerable)
    rpl_themes,
    # Individual theme percentile rankings
    rpl_theme1,   # Socioeconomic Status
    rpl_theme2,   # Household Characteristics
    rpl_theme3,   # Racial & Ethnic Minority Status
    rpl_theme4,   # Housing Type & Transportation
    # Component percentage estimates for deeper analysis
    ep_pov150, ep_unemp, ep_hburd, ep_nohsdp, ep_uninsur,
    ep_age65,  ep_age17, ep_disabl, ep_sngpnt, ep_limeng,
    ep_minrty,
    ep_munit,  ep_mobile, ep_crowd, ep_noveh, ep_groupq
  ) %>%
  # Assign vulnerability tier based on overall SVI quartiles
  mutate(
    svi_tier = case_when(
      rpl_themes <= 0.25 ~ "Low (0-0.25)",
      rpl_themes <= 0.50 ~ "Moderate (0.25-0.50)",
      rpl_themes <= 0.75 ~ "High (0.50-0.75)",
      rpl_themes >  0.75 ~ "Very High (>0.75)",
      TRUE               ~ NA_character_
    ),
    svi_tier = factor(svi_tier, levels = c(
      "Low (0-0.25)", "Moderate (0.25-0.50)",
      "High (0.50-0.75)", "Very High (>0.75)"
    ))
  )

cat("Rows:", nrow(svi_clean), " | Columns:", ncol(svi_clean), "\n")
cat("Missing overall SVI values:", sum(is.na(svi_clean$rpl_themes)), "\n")

```

---

# Exploratory Data Analysis

## Summary Statistics by SVI Theme

```{r eda-summary-stats}

theme_labels <- c(
  rpl_themes = "Overall SVI",
  rpl_theme1 = "Theme 1: Socioeconomic",
  rpl_theme2 = "Theme 2: Household",
  rpl_theme3 = "Theme 3: Minority Status",
  rpl_theme4 = "Theme 4: Housing & Transport"
)

svi_clean %>%
  select(rpl_themes, rpl_theme1, rpl_theme2, rpl_theme3, rpl_theme4) %>%
  pivot_longer(everything(), names_to = "theme", values_to = "rpl") %>%
  group_by(theme) %>%
  summarise(
    n      = sum(!is.na(rpl)),
    mean   = round(mean(rpl,   na.rm = TRUE), 3),
    sd     = round(sd(rpl,     na.rm = TRUE), 3),
    min    = round(min(rpl,    na.rm = TRUE), 3),
    q25    = round(quantile(rpl, 0.25, na.rm = TRUE), 3),
    median = round(median(rpl, na.rm = TRUE), 3),
    q75    = round(quantile(rpl, 0.75, na.rm = TRUE), 3),
    max    = round(max(rpl,    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(theme = recode(theme, !!!theme_labels)) %>%
  kable(
    caption    = "Summary statistics for overall SVI and four theme percentile rankings (county level, 2022)",
    col.names  = c("Theme", "N", "Mean", "SD", "Min", "Q25", "Median", "Q75", "Max")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position   = "left"
  )

```

## Distribution of SVI Scores Across Themes

```{r eda-distributions, fig.cap="Distribution of county-level SVI percentile rankings by theme (2022)"}

svi_clean %>%
  select(rpl_themes, rpl_theme1, rpl_theme2, rpl_theme3, rpl_theme4) %>%
  pivot_longer(everything(), names_to = "theme", values_to = "rpl") %>%
  mutate(theme = recode(theme, !!!theme_labels)) %>%
  ggplot(aes(x = rpl, fill = theme)) +
  geom_histogram(binwidth = 0.05, colour = "white", alpha = 0.85) +
  facet_wrap(~ theme, ncol = 2, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title    = "Distribution of CDC SVI Percentile Rankings by Theme",
    subtitle = "County-level data, United States (2022)",
    x        = "Percentile Ranking (RPL)",
    y        = "Number of Counties",
    caption  = "Source: CDC/ATSDR Social Vulnerability Index 2022"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    strip.text    = element_text(face = "bold")
  )

```

## Vulnerability Tier Breakdown

```{r eda-tier-breakdown, fig.cap="County counts by SVI vulnerability tier"}

tier_counts <- svi_clean %>%
  filter(!is.na(svi_tier)) %>%
  count(svi_tier) %>%
  mutate(pct = n / sum(n))

tier_counts %>%
  kable(
    caption   = "Number and proportion of US counties by SVI vulnerability tier (2022)",
    col.names = c("Vulnerability Tier", "Counties (N)", "Proportion")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "left"
  )

ggplot(tier_counts, aes(x = svi_tier, y = n, fill = svi_tier)) +
  geom_col(colour = "white", width = 0.7) +
  geom_text(aes(label = paste0(n, "\n(", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_brewer(palette = "RdYlGn", direction = -1, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "US Counties by SVI Vulnerability Tier (2022)",
    subtitle = "Tiers defined by quartiles of the overall RPL_THEMES score",
    x        = "Vulnerability Tier",
    y        = "Number of Counties",
    caption  = "Source: CDC/ATSDR Social Vulnerability Index 2022"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40")
  )

```

## State-Level Mean SVI Rankings

```{r eda-state-means, fig.height=9, fig.cap="Mean overall SVI score by state (2022)"}

state_means <- svi_clean %>%
  filter(!is.na(rpl_themes)) %>%
  group_by(state) %>%
  summarise(
    n_counties   = n(),
    mean_svi     = mean(rpl_themes, na.rm = TRUE),
    mean_theme1  = mean(rpl_theme1, na.rm = TRUE),
    mean_theme2  = mean(rpl_theme2, na.rm = TRUE),
    mean_theme3  = mean(rpl_theme3, na.rm = TRUE),
    mean_theme4  = mean(rpl_theme4, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_svi))

ggplot(state_means, aes(x = mean_svi,
                        y = reorder(state, mean_svi),
                        fill = mean_svi)) +
  geom_col(colour = "white", width = 0.75) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Mean SVI") +
  scale_x_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.03))) +
  labs(
    title    = "Mean Overall CDC SVI Score by State (2022)",
    subtitle = "Average of county-level RPL_THEMES (0 = least, 1 = most vulnerable)",
    x        = "Mean SVI Percentile Ranking",
    y        = NULL,
    caption  = "Source: CDC/ATSDR Social Vulnerability Index 2022"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    legend.position = "right"
  )

```

---

# Spatial Analysis

## Choropleth Map — Overall SVI

```{r spatial-load-counties}

# Download US county boundaries (2022, no water areas)
counties_sf <- counties(cb = TRUE, resolution = "20m", year = 2022) %>%
  shift_geometry() %>%
  clean_names() %>%
  mutate(fips = paste0(statefp, countyfp))

# Join SVI data to spatial boundaries
svi_sf <- counties_sf %>%
  left_join(svi_clean, by = "fips") %>%
  filter(!is.na(rpl_themes))

```

```{r spatial-map-overall, fig.width=12, fig.height=7, fig.cap="Choropleth map of overall CDC SVI score by county, contiguous US (2022)"}

ggplot(svi_sf) +
  geom_sf(aes(fill = rpl_themes), colour = NA) +
  scale_fill_viridis_c(
    option   = "magma",
    direction = -1,
    name     = "SVI\nPercentile",
    labels   = percent_format(accuracy = 1),
    na.value = "grey80"
  ) +
  labs(
    title    = "CDC Social Vulnerability Index — Overall Score (2022)",
    subtitle = "County-level RPL_THEMES: 0 = least vulnerable, 1 = most vulnerable",
    caption  = "Source: CDC/ATSDR SVI 2022 | Boundaries: US Census Bureau TIGER/Line"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    plot.caption  = element_text(hjust = 0.5, color = "grey50", size = 9),
    legend.position = "right"
  )

```

## Choropleth Maps — Four SVI Themes

```{r spatial-map-themes, fig.width=14, fig.height=10, fig.cap="Choropleth maps by CDC SVI theme (county level, 2022)"}

theme_var_labels <- c(
  rpl_theme1 = "Theme 1: Socioeconomic Status",
  rpl_theme2 = "Theme 2: Household Characteristics",
  rpl_theme3 = "Theme 3: Racial & Ethnic Minority",
  rpl_theme4 = "Theme 4: Housing & Transportation"
)

plot_theme_map <- function(var, label) {
  ggplot(svi_sf) +
    geom_sf(aes(fill = .data[[var]]), colour = NA) +
    scale_fill_viridis_c(
      option    = "plasma",
      direction = -1,
      name      = "RPL",
      labels    = percent_format(accuracy = 1),
      na.value  = "grey80"
    ) +
    labs(title = label) +
    theme_void(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 11),
      legend.position = "right",
      legend.key.height = unit(0.8, "cm")
    )
}

map_plots <- imap(theme_var_labels, plot_theme_map)

grid.arrange(grobs = map_plots, ncol = 2,
             top = "CDC SVI Percentile Rankings by Theme — US Counties (2022)")

```

---

# Correlation Analysis

## Pairwise Correlations Between SVI Themes

```{r correlation-matrix, fig.cap="Pairwise scatter plots of SVI theme percentile rankings (county level, 2022)"}

svi_cor_data <- svi_clean %>%
  select(
    `Overall SVI`     = rpl_themes,
    `Theme 1\nSocioeconomic` = rpl_theme1,
    `Theme 2\nHousehold`     = rpl_theme2,
    `Theme 3\nMinority`      = rpl_theme3,
    `Theme 4\nHousing`       = rpl_theme4
  ) %>%
  drop_na()

cor_matrix <- cor(svi_cor_data, use = "complete.obs") %>%
  round(3)

cor_matrix %>%
  kable(caption = "Pearson correlation matrix of SVI overall score and theme rankings") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "left"
  )

```

## Scatter — Socioeconomic vs. Minority Status by Vulnerability Tier

```{r scatter-theme1-theme3, fig.cap="County-level SVI Theme 1 (Socioeconomic) vs. Theme 3 (Minority Status) by vulnerability tier"}

svi_clean %>%
  filter(!is.na(rpl_theme1), !is.na(rpl_theme3), !is.na(svi_tier)) %>%
  ggplot(aes(x = rpl_theme1, y = rpl_theme3, colour = svi_tier)) +
  geom_point(alpha = 0.45, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, colour = "black",
              linetype = "dashed") +
  scale_colour_brewer(palette = "RdYlGn", direction = -1,
                      name = "Vulnerability Tier") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title    = "Socioeconomic Status vs. Racial & Ethnic Minority Status",
    subtitle = "County-level SVI theme percentile rankings (2022)",
    x        = "Theme 1 RPL: Socioeconomic Status",
    y        = "Theme 3 RPL: Racial & Ethnic Minority",
    caption  = "Source: CDC/ATSDR Social Vulnerability Index 2022"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    legend.position = "bottom"
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3)))

```

## Top 25 Most Vulnerable Counties

```{r top-vulnerable-counties}

svi_clean %>%
  filter(!is.na(rpl_themes)) %>%
  slice_max(rpl_themes, n = 25) %>%
  select(
    State          = state,
    County         = county,
    Population     = e_totpop,
    `Overall SVI`  = rpl_themes,
    `Theme 1`      = rpl_theme1,
    `Theme 2`      = rpl_theme2,
    `Theme 3`      = rpl_theme3,
    `Theme 4`      = rpl_theme4
  ) %>%
  mutate(across(where(is.numeric) & !Population, ~ percent(., accuracy = 0.1)),
         Population = scales::comma(Population)) %>%
  kable(caption = "Top 25 most socially vulnerable US counties by overall SVI score (2022)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "left"
  ) %>%
  column_spec(4, bold = TRUE, color = "white",
              background = spec_color(
                as.numeric(gsub("%", "", svi_clean %>%
                  filter(!is.na(rpl_themes)) %>%
                  slice_max(rpl_themes, n = 25) %>%
                  pull(rpl_themes))),
                end = 0.9, option = "magma", direction = -1))

```

---

# Literature Cited

```{r literature-cited, echo=FALSE, include=FALSE}

references_path <- here("Scripts", "Docs", "References")

# Generate citations for all loaded packages
knitr::write_bib(.packages(), here(references_path, "references.bib"))

# Save session information
session_info_obj <- sessioninfo::session_info()

write.csv(session_info_obj$platform %>% as_tibble(),
          here(references_path, "session_info.csv"))
write.csv(session_info_obj$packages %>% as_tibble(),
          here(references_path, "session_info_packages.csv"))

```
